---
title: "clustOpt: Quickstart Guide"
output: rmarkdown::html_vignette
date: 'Compiled: `r Sys.Date()`'
vignette: >
  %\VignetteIndexEntry{clustOpt: Quickstart Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  fig.height = 5,
  fig.width = 8,
  cache = TRUE,
  cache.lazy = FALSE
)
```

```{r setup, message=FALSE}
library(Seurat)
# Only v5 Seurat Assays are supported
options(Seurat.object.assay.version = "v5")
# Increase future globals limit for large datasets
options(future.globals.maxSize = 2e9)  # 2GB limit
library(dplyr)
library(tidyr)
library(ggplot2)
library(clustOpt)
library(future)
set.seed(42)
```

Read in the tutorial dataset. This data was generated from the [Asian Immune Diversity Atlas](https://doi.org/10.1101/2024.06.30.601119) (data freeze v1). It contains only B cells, Monocytes, and NK cells from 10 donors which have been sketched to a total of 1000 cells using the leverage score based method of Seurat's SketchData function.

```{r load_data}
input <- readRDS(
  system.file(
    "extdata",
    "1000_cell_sketch_10_donors_3_celltypes_AIDA.rds",
    package = "clustOpt"
  )
)
```


```{r summarize}
input@meta.data |>
  summarize(
    Donors = n_distinct(donor_id),
    Celltypes = n_distinct(broad_cell_type),
    `Cell Subtypes` = n_distinct(author_cell_type),
    Ethnicities = n_distinct(Ethnicity_Selfreported),
    Countries = n_distinct(Country)
  ) |>
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Count")
```


```{r process}
input <- input |>
  SCTransform(verbose = FALSE) |>
  RunPCA(verbose = FALSE)
```

```{r elbowplot}
ElbowPlot(input, ndims = 50)
```

```{r UMAP}
input <- RunUMAP(input, dims = 1:20, verbose = FALSE)

DimPlot(input,
  group.by = "donor_id",
  pt.size = .8
) +
  coord_fixed()
```

```{r}
DimPlot(input,
  group.by = "broad_cell_type",
  pt.size = .8
) +
  coord_fixed()
```

```{r}
DimPlot(input,
  group.by = "author_cell_type",
  pt.size = .8
) +
  coord_fixed()
```

## How clustOpt Works

clustOpt uses subject-wise leave-one-out cross-validation to find clustering resolutions that generalize across biological replicates:

1. **PC Splitting**: Principal components are split into two independent sets (odd: PC1, PC3, PC5... and even: PC2, PC4, PC6...)
2. **Training**: For each held-out subject, a Random Forest is trained on one PC set to predict cluster assignments
3. **Evaluation**: Clustering quality metrics are calculated on the opposite PC set

This separation ensures that the evaluation is performed on data independent from what was used for training, preventing overfitting and providing a more honest assessment of clustering reproducibility.

## Running clustOpt

Subject-wise cross validation is parallelized with `future` and `future.batchtools` can be used for HPC job schedulers. For details on which plan to use for your setup, see the future package documentation. The code below requires a large amount of RAM (32 GB recommended) and 10 cores to run in about 15 minutes.

```{r clustOpt, warning=FALSE, cache=TRUE}
plan("multisession", workers = 11)

cv_results <- clust_opt(input,
  subject_ids = "donor_id",
  ndim = 20,
  verbose = 1,
  train_with = "even"
) |>
  progressr::with_progress() # Optional, just adds progress bar

# Shut down stray workers
plan("sequential")
```

## Plotting the silhouette score distributions

```{r plotting}
plots <- create_sil_plots(cv_results |> drop_na())

plots[[1]]
plots[[2]]
plots[[3]]
plots[[4]]
```

The silhouette score distribution shows how consistently clusters are separated across subjects at each resolution. Key patterns to look for:

- **Declining scores at higher resolutions** indicate over-clustering, where cells are forced into clusters that don't represent distinct populations
- **Low variance across subjects** (narrow boxes) indicates reproducible clustering
- **Local maxima** may indicate resolutions that capture natural biological structure

The optimal resolution typically shows high median silhouette scores with low variance across subjects, suggesting that the clustering captures cell populations that are consistently distinguishable regardless of which subjects are used for training.

## Additional Quality Metrics

> **Note:** We currently recommend **silhouette scores** for resolution selection. The additional metrics below are still undergoing benchmarking across different experimental settings.

`clust_opt()` also computes the following at each resolution:

| Metric | Measures | Better |
|--------|----------|--------|
| **avg_width** | Silhouette score: cluster separation (recommended) | Higher |
| **KLdivergence** | KL divergence of cluster proportions (experimental) | Lower |
| **Hellinger** | Hellinger distance of cluster proportions (experimental) | Lower |
| **modularity** | Graph modularity of cluster assignments (experimental) | Higher |

```{r metrics_overview}
head(cv_results)
```

```{r plot_metrics, fig.height=10}
cv_results |>
  drop_na() |>
  pivot_longer(cols = c(KLdivergence, Hellinger, modularity),
               names_to = "metric", values_to = "value") |>
  ggplot(aes(x = as.factor(resolution), y = value)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  theme_bw() +
  labs(x = "Resolution", y = NULL)
```

## Suggesting a Resolution

`suggest_resolution()` provides two approaches for identifying optimal resolutions:

- **`"rank"`** (default): Ranks all tested resolutions by aggregating their median metric values. Works with any number of subjects.

- **`"local_optima"`**: Filters to resolutions with reproducible clustering (low Hellinger distance across subjects), then ranks the filtered set using both rank aggregation and curvature-based detection of local optima in the metric curves. Recommended when you have >=10 subjects, as the Hellinger filter thresholds scale automatically with the number of subjects.

`plot_rank_metrics()` shows the per-metric ranks and `plot_mean_rank()` shows the aggregate.

```{r suggest_resolution}
rankings <- suggest_resolution(cv_results)
rankings
```

```{r plot_rankings}
plot_rank_metrics(rankings)
```

```{r plot_mean_rank}
plot_mean_rank(rankings)
```

This tutorial uses the default `"rank"` method. The example dataset contains only 3 well-separated celltypes, so a simple rank aggregation is sufficient. For real-world datasets with a denser resolution grid and more ambiguous cluster structure, `method = "local_optima"` can better identify resolutions that sit at natural inflection points in the metric curves. See `?suggest_resolution` for details.

## Comparison to the cell annotations

```{r cluster}
input <- FindNeighbors(input, dims = 1:20, verbose = FALSE)
input <- FindClusters(input, resolution = 0.04)

DimPlot(input,
  group.by = "seurat_clusters",
  pt.size = .8
)
```

At the reproducible resolution there are 3 clusters, corresponding to the 3 celltypes. Use `adjusted_rand_index()` to compare any two metadata columns:

```{r}
adjusted_rand_index(input,
  meta1 = "seurat_clusters",
  meta2 = "broad_cell_type"
)
adjusted_rand_index(input,
  meta1 = "seurat_clusters",
  meta2 = "author_cell_type"
)
```

## Larger Datasets

Due to the resource requirements of clustOpt, we recommend running it in Rscripts with the `future` plan set to "multicore" instead of "multisession" and if possible using `future.batchtools`. Additionally we provide a wrapper around `Seurat`'s SketchData using the leverage score method to reduce the size of the input data to clustOpt.

```{r}
input <- leverage_sketch(input, sketch_size = 100)
input
```

