---
title: "clustOpt: Quickstart Guide"
output: rmarkdown::html_vignette
date: 'Compiled: `r Sys.Date()`'
vignette: >
  %\VignetteIndexEntry{clustOpt: Quickstart Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  fig.height = 5,
  fig.width = 8,
  cache = TRUE,
  cache.lazy = FALSE
)
```

```{r setup, message=FALSE}
library(Seurat)
# Only v5 Seurat Assays are supported
options(Seurat.object.assay.version = "v5")
# Increase future globals limit for large datasets
options(future.globals.maxSize = 2e9)  # 2GB limit
library(dplyr)
library(tidyr)
library(ggplot2)
library(clustOpt)
library(future)
set.seed(42)
```

Read in the tutorial dataset. This data was generated from the [Asian Immune Diversity Atlas](https://doi.org/10.1101/2024.06.30.601119) (data freeze v1). It contains only B cells, Monocytes, and NK cells from 10 donors which have been sketched to a total of 1000 cells using the leverage score based method of Seurat's SketchData function.

```{r load_data}
input <- readRDS(
  system.file(
    "extdata",
    "1000_cell_sketch_10_donors_3_celltypes_AIDA.rds",
    package = "clustOpt"
  )
)
```


```{r summarize}
input@meta.data |>
  summarize(
    Donors = n_distinct(donor_id),
    Celltypes = n_distinct(broad_cell_type),
    `Cell Subtypes` = n_distinct(author_cell_type),
    Ethnicities = n_distinct(Ethnicity_Selfreported),
    Countries = n_distinct(Country)
  ) |>
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Count")
```


```{r process}
input <- input |>
  SCTransform(verbose = FALSE) |>
  RunPCA(verbose = FALSE)
```

```{r elbowplot}
ElbowPlot(input, ndims = 50)
```

```{r UMAP}
input <- RunUMAP(input, dims = 1:20, verbose = FALSE)

DimPlot(input,
  group.by = "donor_id",
  pt.size = .8
) +
  coord_fixed()
```

```{r}
DimPlot(input,
  group.by = "broad_cell_type",
  pt.size = .8
) +
  coord_fixed()
```

```{r}
DimPlot(input,
  group.by = "author_cell_type",
  pt.size = .8
) +
  coord_fixed()
```

Subject-wise cross validation is parallelized with `future` and `future.batchtools` can be used for HPC job schedulers. For details on which plan to use for your setup, see the future package documentation. The code below requires a large amount of RAM (32 GB recommended) and 10 cores to run in about 20 minutes. The repeated messages about "SeuratObject" are a side effect of using the future plan "multisession" and should not occur with other future plans.

```{r clustOpt, warning=FALSE, cache=TRUE}
plan("multisession", workers = 10)

cv_results <- clust_opt(input,
  subject_ids = "donor_id",
  ndim = 20,
  verbose = FALSE,
  train_with = "even"
) |>
  progressr::with_progress() # Optional, just adds progress bar

# Shut down stray workers
plan("sequential")
```

## Plotting the silhouette score distributions

```{r plotting}
plots <- create_sil_plots(cv_results |> drop_na())

plots[[1]]
plots[[2]]
plots[[3]]
plots[[4]]
```

The median silhouette scores across all cells start to fall off at 0.2, indicating that this clustering resolution is the most reproducible across all biological replicates. It is also possible to see a local maximum which would indicate a reproducible resolution.

## Additional Quality Metrics

The `clust_opt()` function calculates several metrics beyond silhouette scores to help evaluate clustering quality. Let's examine the full output:

```{r metrics_overview}
# View all available metrics
colnames(cv_results)

# Preview the data
head(cv_results)
```

### Understanding the Metrics

- **avg_width / cluster_median_widths**: Silhouette scores measuring cluster separation
- **mse / mad**: Mean Squared Error and Mean Absolute Deviation measure within-cluster compactness
- **KLdivergence / Hellinger**: Measure similarity between training and predicted cluster distributions
- **modularity**: Graph-based metric measuring clustering coherence

### Within-Cluster Compactness (MSE & MAD)

Mean Squared Error (MSE) and Mean Absolute Deviation (MAD) measure how tightly cells are grouped around their cluster centroids. Lower values indicate more compact clusters. These metrics will generally decrease as resolution increases (more clusters = tighter groupings), so they are most useful for comparing resolutions with similar silhouette scores.

```{r plot_mse}
cv_results |>
  drop_na() |>
  ggplot(aes(x = as.factor(resolution), y = mse)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = "Mean Squared Error by Resolution",
    x = "Resolution",
    y = "MSE"
  )
```

```{r plot_mad}
cv_results |>
  drop_na() |>
  ggplot(aes(x = as.factor(resolution), y = mad)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = "Mean Absolute Deviation by Resolution",
    x = "Resolution",
    y = "MAD"
  )
```

### Distribution Similarity (KL Divergence & Hellinger Distance)

These metrics measure how similar the cluster proportions are between the training data and the predicted test data. Lower values indicate that the clustering is more reproducible across subjects. KL Divergence is asymmetric and can be more sensitive to rare clusters, while Hellinger Distance is symmetric and bounded between 0 and 1.

```{r plot_kl}
cv_results |>
  drop_na() |>
  ggplot(aes(x = as.factor(resolution), y = KLdivergence)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = "KL Divergence by Resolution",
    x = "Resolution",
    y = "KL Divergence"
  )
```

```{r plot_hellinger}
cv_results |>
  drop_na() |>
  ggplot(aes(x = as.factor(resolution), y = Hellinger)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = "Hellinger Distance by Resolution",
    x = "Resolution",
    y = "Hellinger Distance"
  )
```

### Graph-Based Clustering Quality (Modularity)

Modularity measures how well the predicted clusters partition a k-nearest neighbor graph of the test data. Higher values indicate that cells within clusters are more connected to each other than to cells in other clusters. This provides a complementary view of cluster quality based on local neighborhood structure rather than global distances.

```{r plot_modularity}
cv_results |>
  drop_na() |>
  ggplot(aes(x = as.factor(resolution), y = modularity)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = "Modularity by Resolution",
    x = "Resolution",
    y = "Modularity"
  )
```

### Interpreting Multiple Metrics

When selecting an optimal resolution, consider multiple metrics together:

1. **Silhouette scores** (primary): Look for the resolution where scores begin to decline
2. **KL Divergence / Hellinger**: Lower values indicate the clustering is reproducible across subjects
3. **Modularity**: Higher values suggest better-defined cluster boundaries in the kNN graph
4. **MSE/MAD**: Lower is better, but will always decrease with more clusters; useful for comparing similar resolutions

## Comparison to the cell annotations

```{r cluster}
input <- FindNeighbors(input, dims = 1:20, verbose = FALSE)
input <- FindClusters(input, resolution = 0.2)

DimPlot(input,
  group.by = "seurat_clusters",
  pt.size = .8
)
```

We can see that at the reproducible resolution there are 4 clusters, with the naive B cells being distinct enough to be separated into their own cluster based on the ability to detect them across the CV folds. In other words the naive B cells are present in high enough amounts across the subjects and transcriptomically distinct enough to be consistently detected.

For convenience we provide a function to calculate the adjusted rand index for any 2 metadata columns in a seurat object so that the clustOpt resolution parameter can be compared to clusters derived from other methods.

```{r}
adjusted_rand_index(input,
  meta1 = "seurat_clusters",
  meta2 = "broad_cell_type"
)
adjusted_rand_index(input,
  meta1 = "seurat_clusters",
  meta2 = "author_cell_type"
)
```

## Larger Datasets

Due to the resource requirements of clustOpt, we recommend running it in Rscripts with the `future` plan set to "multicore" instead of "multisession" and if possible using `future.batchtools`. Additionally we provide a wrapper around `Seurat`'s SketchData using the leverage score method to reduce the size of the input data to clustOpt.

```{r}
input <- leverage_sketch(input, sketch_size = 100)
input
```

